---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fastGFPCA

<!-- badges: start -->
<!-- badges: end -->

fast generalized functional principal components analysis

* Authors: [Julia Wrobel](http://juliawrobel.com), Ciprian Crainiceanu, Andrew Leroux
* License: [MIT](https://opensource.org/licenses/MIT). See the [LICENSE](LICENSE) file for details
* Version: 0.9


## Installation

You can install the development version of fastGFPCA from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("julia-wrobel/fastGFPCA")


# build vignette with installation
devtools::install_github("julia-wrobel/fastGFPCA", build_vignettes = TRUE)
```

## Example

This example performs GFPCA on simulated binary data. More details on the use of the package can be found in the vignette.

```{r example}
library(fastGFPCA)

# simulate data
set.seed(10001)
df_gfpca <- sim_gfpca(N = 50, J = 500, case = 1)$df_gfpca

# run gfpca
gfpca_mod <- fast_gfpca(df_gfpca, overlap = TRUE, binwidth = 10, npc = 4, family = "binomial")
```



